Parameter Recovery Validation

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): remove references to hierarchical temporal prior
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

- **Time coordinates**: Same hierarchical latent $t^*_j$ structure as previous model
- **Dynamics**: Piecewise constant transcription with three phases (off, on, return-to-off)
- **Parameters**: $\gamma^*_i$, $R_{\text{on},i}$ (fold-change), $t^*_{0,\text{on}i}$ (onset), $\delta^*_i$ (duration), plus hierarchical time structure
- **Key innovation**: Fixed reference state ($\alpha^*_{\text{off}i} = 1.0$) eliminates parameter redundancy
- **Enhanced patterns**: Supports pre-activation scenarios with negative onset times
- **Analytical solutions**: Closed-form solutions for all three phases, including $\gamma^*_i = 1$ special case
- **Biological realism**: Steady-state initial conditions and rich expression pattern representation
- **Observation Model**: Same Poisson scaling with characteristic scales and capture efficiency
- **Likelihood**: Poisson

@@

- $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ (early activation within process duration)
- $\delta^*_i < 0.4 \mathbb{E}[T_{\max}]$ (short activation duration relative to process)
- $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)

@@

- $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ (early activation within process duration)
- $\delta^*_i > 0.5 \mathbb{E}[T_{\max}]$ (long activation duration relative to process)
- $R_{\text{on},i} > 2.0$ (strong fold-change during activation)

@@

```{python}
#| label: fig-piecewise-activation-pgm
#| code-fold: true
#| fig-cap: Probabilistic graphical model for dimensionless analytical dynamics with hierarchical latent time coordinates and piecewise activation.

@@

- **$t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around $0.3 \mathbb{E}[T_{\max}]$ to expect activation at 30% of process duration. **Negative values** enable pre-activation scenarios where genes were already activated before observation began.

- **$\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$**: Absolute activation duration, typically centered around $0.4 \mathbb{E}[T_{\max}]$ to expect 40% duration activation. This enables both transient (short duration) and sustained (long duration) activation patterns.

@@

- **Pre-activation** ($t^*_{0,\text{on}i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
- **Transient** ($0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ and $\delta^*_i < 0.4 \mathbb{E}[T_{\max}]$): Balanced probability for observable complete cycles
- **Sustained** ($0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ and $\delta^*_i > 0.5 \mathbb{E}[T_{\max}]$): Balanced probability for long-duration activation

@@

2. **Temporal coordination**: Does the hierarchical time structure produce biologically plausible cell orderings and temporal relationships?

@@

1. **Parameter uncertainty**: $\text{Var}[\theta | y_{obs}]$ from the posterior distribution over all model parameters
2. **Temporal uncertainty**: Uncertainty in latent time coordinates $t^*_j$ and their hierarchical structure
3. **Kinetic uncertainty**: Uncertainty in gene-specific kinetic parameters $\gamma^*_i$ and fold-changes $R_{\text{on},i}$
4. **Concentration uncertainty**: Uncertainty in predicted latent concentrations $u^*_{ij}(\theta), s^*_{ij}(\theta)$ arising from parameter uncertainty
5. **Observation uncertainty**: Poisson noise in final count observations

@@

- **Pre-activation scenarios**: Models genes already activated when observation begins
- **Flexible onset timing**: Normal prior allows negative onset times naturally
- **Biological realism**: Captures realistic experimental scenarios better
- **Maintains population structure**: Hierarchical time modeling preserved

@@

- **Dimensionless parameters**: Universal interpretation across experimental conditions
- **Temporal coordination**: Hierarchical time structure captures population-level timing
- **Gene-specific dynamics**: Each gene can follow distinct activation pattern

>>>>

- **Time coordinates**: Independent latent $t^*_j$ coordinates sampled uniformly within process duration $T_{\max}$
- **Dynamics**: Piecewise constant transcription with three phases (off, on, return-to-off)
- **Parameters**: $\gamma^*_i$, $R_{\text{on},i}$ (fold-change), $t^*_{0,\text{on}i}$ (onset), $\delta^*_i$ (duration), with independent temporal structure
- **Key innovation**: Fixed reference state ($\alpha^*_{\text{off}i} = 1.0$) eliminates parameter redundancy
- **Enhanced patterns**: Supports pre-activation scenarios with negative onset times
- **Analytical solutions**: Closed-form solutions for all three phases, including $\gamma^*_i = 1$ special case
- **Biological realism**: Steady-state initial conditions and rich expression pattern representation
- **Observation Model**: Same Poisson scaling with characteristic scales and capture efficiency
- **Likelihood**: Poisson

@@

- $0 < t^*_{0,\text{on}i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
- $\delta^*_i < 2.0$ (short activation duration relative to process, assuming conservative prior Gamma(5,1))
- $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)

@@

- $0 < t^*_{0,\text{on}i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
- $\delta^*_i > 2.5$ (long activation duration relative to process, assuming conservative prior Gamma(5,1))
- $R_{\text{on},i} > 2.0$ (strong fold-change during activation)

@@

```{python}
#| label: fig-piecewise-activation-pgm
#| code-fold: true
#| fig-cap: Probabilistic graphical model for dimensionless analytical dynamics with independent latent time coordinates and piecewise activation.

@@

- **$t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around 1.5 to expect activation at 30% of process duration (assuming conservative prior Gamma(5,1) with mean 5). **Negative values** enable pre-activation scenarios where genes were already activated before observation began.

- **$\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$**: Absolute activation duration, typically centered around 2.0 to expect 40% duration activation (assuming conservative prior Gamma(5,1) with mean 5). This enables both transient (short duration) and sustained (long duration) activation patterns.

@@

- **Pre-activation** ($t^*_{0,\text{on}i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
- **Transient** ($0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i < 2.0$): Balanced probability for observable complete cycles (assuming conservative prior Gamma(5,1))
- **Sustained** ($0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i > 2.5$): Balanced probability for long-duration activation (assuming conservative prior Gamma(5,1))

@@

2. **Temporal coordination**: Does the independent time structure produce biologically plausible cell orderings and temporal relationships?

@@

1. **Parameter uncertainty**: $\text{Var}[\theta | y_{obs}]$ from the posterior distribution over all model parameters
2. **Temporal uncertainty**: Uncertainty in latent time coordinates $t^*_j$ and process duration $T_{\max}$
3. **Kinetic uncertainty**: Uncertainty in gene-specific kinetic parameters $\gamma^*_i$ and fold-changes $R_{\text{on},i}$
4. **Concentration uncertainty**: Uncertainty in predicted latent concentrations $u^*_{ij}(\theta), s^*_{ij}(\theta)$ arising from parameter uncertainty
5. **Observation uncertainty**: Poisson noise in final count observations

@@

- **Pre-activation scenarios**: Models genes already activated when observation begins
- **Flexible onset timing**: Normal prior allows negative onset times naturally
- **Biological realism**: Captures realistic experimental scenarios better
- **Maintains temporal structure**: Independent time modeling with process duration constraints

@@

- **Dimensionless parameters**: Universal interpretation across experimental conditions
- **Temporal coordination**: Independent time structure with process duration constraints captures temporal relationships
- **Gene-specific dynamics**: Each gene can follow distinct activation pattern
